A Computational Account of Default-Mode Function\\ by Control Theory and Reinforcement Learning

### 2016-10-02 paper/paper.tex
thoughts in a train

\section{Reinforcement-Learning as a computational model for DMN function}
The main idea is to show that the DMN has biological substrates (gradient
computation, backprop, experience replay, etc.) for all the
major components of an efficient RL algorithm, and all these components
integrate in a plausible way.
\subsection{Markov Decision Processes}
In a general reinforcement-learning (RL) framework, an \textit{agent}
interacts with an  environment in a trial-by-error fashion by taking
actions / decisions $a$, which trigger changes in the
\textit{state} of the environment
$s \rightarrow s'$, accompanied with a real-valued \textit{reward}
$r = r(s, a)$ (or regret!, since it can be negative) collected by the
agent. This reward should be thought
of as an instantaneous satisfaction accompanying the execution of
an action. Here the subscript $t$ denotes time, and can be discrete or
continuous.
The environment $\mathcal E$ is generally taken as stochastic,
is partially observed and 
has some kind of memory.
A popular abstract model for the scenario are so-called \textit{Markov Decision Processes (MDPs)}.
In RL, MDPs are the mathematical instantiation of the environment model,
which generally can incorporate action probabilities (policy matrix)
and/or value function given a state.
(Silver: The environment can be fully defined by state space and
transition probability matrix.
Across lifetime, the MDP naturally incorporates always better
models of the environment due to \points.)
This is simply a quintuplet $(\mathcal S, \mathcal A, r, \rho, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is any function of the history.
The state exhibits the Markov property
$\mathbb{P}(S_{t+1}|S_{t})=\mathbb{P}(S_{t+1}|S_{1}, \dots, S_{t})$.
Consequently, the future states are independent of past states and only depend on the present.
% \item $\mathcal S$ is the set of states, e.g $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $\mathcal A$ is the set of actions, e.g $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$, so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts-off in state $s$.
  \item $\rho : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto \rho(s'|s,a)$, the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such transitions be Markovian.
\end{itemize}

@@

\subsection{Correspondence between DMN function and RL framework}
\textbf{XXX Put the diagrams and process charts here.}
\begin{itemize}
  \item The policy matrix represents the repertoire of possible actions
  on the world given a current state. It encodes the probabilites of
  choosing actions to be executed in a certain situation.
  \item Given a state, the value function represents the estimated utility of
  being in that state, which is subjective to the agent. 
  For state s there is set of actions a.
  It provides a quantitative answer to the question:
  \"What state is desirable for a specific agent.\"
  For all actions being in one of the states s,
  the Q value function provides the subjective utility
  of executing a specific action.
  It provides a quantitative answer to the question:
  \"Which is the best action to take in the current situation.\"
  \item The DMN may subserve
  constant exploration of possible future actions and their
  cumulative outcomes. Implicit computation of future choices 
  provides an explanation for purposeful mind-wandering.
  \item hippocampus -> provides perturbed action-transition-state-reward samples
  as a block of "imagined", "hypothesized", "recalled" experience;
  the small variations in these experience blocks allow searching
  a larger space of parameters and possible experiences.
  In the absence of environmental input and feedback
  (e.g., mind-wandering or sleep) pseudo-experiencing possible
  future scenarios and action outcomes.
  \item
  \textit{Inference} in the human brain reduces to generalization of
  policy and value adaptations from sampled experiences to
  successful action choices and reward predictions in future states.

>>>>

\section{Reinforcement-Learning as a computational model for DMN function}
The main idea is to show that the DMN has biological substrates (gradient
computation, backprop, experience replay, etc.) for all the
major components of an efficient RL algorithm, and all these components
integrate in a plausible way.
\subsection{Markov Decision Processes}
In a general reinforcement-learning (RL) framework, an \textit{agent}
interacts with an  environment in a trial-by-error fashion by taking
actions / decisions $a$, which trigger changes in the
\textit{state} of the environment
$s \rightarrow s'$, accompanied with a real-valued \textit{reward}
$r = r(s, a)$ (or regret!, since it can be negative) collected by the
agent. In this view, the environmental is partially controlled by
the action of the agent.
This reward should be thought
of as an instantaneous satisfaction accompanying the execution of
an action. Here the subscript $t$ denotes time, and can be discrete or
continuous.
The environment $\mathcal E$ is generally taken as stochastic,
is partially observed and 
has some kind of memory.
A popular abstract model for the scenario are so-called \textit{Markov Decision Processes (MDPs)}.
In RL, MDPs are the mathematical instantiation of the environment model,
which generally can incorporate action probabilities (policy matrix)
and/or value function given a state.
(Silver: The environment can be fully defined by state space and
transition probability matrix.
Across lifetime, the MDP naturally incorporates always better
models of the environment due to \points.)
This is simply a quintuplet $(\mathcal S, \mathcal A, r, \rho, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is any function of the history.
The state exhibits the Markov property
$\mathbb{P}(S_{t+1}|S_{t})=\mathbb{P}(S_{t+1}|S_{1}, \dots, S_{t})$.
Consequently, the future states are independent of past states and only depend on the present.
% \item $\mathcal S$ is the set of states, e.g $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $\mathcal A$ is the set of actions, e.g $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$, so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts-off in state $s$.
  \item $\rho : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto \rho(s'|s,a)$, the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such transitions be Markovian.
\end{itemize}

@@

\subsection{Correspondence between DMN function and RL framework}
\textbf{XXX Put the diagrams and process charts here.}
\begin{itemize}
  \item The policy matrix represents the repertoire of possible actions
  on the world given a current state. It encodes the probabilites of
  choosing actions to be executed in a certain situation.
  \item Given a state, the value function represents the estimated utility of
  being in that state, which is subjective to the agent. 
  For state s there is set of actions a.
  It provides a quantitative answer to the question:
  \"What state is desirable for a specific agent.\"
  For all actions being in one of the states s,
  the Q value function provides the subjective utility
  of executing a specific action.
  It provides a quantitative answer to the question:
  \"Which is the best action to take in the current situation.\"
  \item The DMN may subserve
  constant exploration of possible future actions and their
  cumulative outcomes. Implicit computation of future choices 
  provides an explanation for purposeful mind-wandering.
  \item hippocampus -> provides perturbed action-transition-state-reward samples
  as a block of "imagined", "hypothesized", "recalled" experience;
  the small variations in these experience blocks allow searching
  a larger space of parameters and possible experiences.
  In the absence of environmental input and feedback
  (e.g., mind-wandering or sleep) pseudo-experiencing possible
  future scenarios and action outcomes.
  \item
  \textit{Inference} in the human brain reduces to generalization of
  policy and value adaptations from sampled experiences to
  successful action choices and reward predictions in future states.
  \item In the RL view, plasticity in the DMN arises naturally.
  If an agent behaving optimally in a given environment moves
  to novel, yet unexperienced environments, reward prediction errors will
  massively increase. This will lead to adaptation of the policy matrix
  and value function until the system converges to a
  new steady-state of optimal action decisions given certain states.
